---
title: "Practical 3"
author: "Rifat Fariha B00937648"
date: "2024-06-08"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As we found for natural language processing, R is a little bit limited when it comes to deep neural networks. Unfortunately, these networks are the current gold-standard methods for medical image analysis. The ML/DL R packages that do exist provide interfaces (APIs) to libraries built in other languages like C/C++ and Java. This means there are a few “non-reproducible” installation and data gathering steps to run first. It also means there are a few “compromises” to get a practical that will A) be runnable B) actually complete within the allotted time and C) even vaguely works.
Doing this on a full real dataset would likely follow a similar process (admittedly with more intensive training and more expressive architectures). Google colab running an R kernel hasn't proven a very effective alternative thus far either.

# Install packages

In this practical, we will be performing image classification on real medical images using the Torch deep learning library and several packages to manage and preprocess the images. We will utilize the following packages for our task:

`gridExtra`: This package provides functions for arranging multiple plots in a grid layout, allowing us to visualize and compare our results effectively.

`jpeg`: This package enables us to read and write JPEG images, which is a commonly used image format.

`imager`: This package offers a range of image processing functionalities, such as loading, saving, resizing, and transforming images.

`magick`: This package provides an interface to the ImageMagick library, allowing us to manipulate and process images using a wide variety of operations.

To ensure that the required packages are available, we can install them using the `install.packages()` function. Additionally, some packages may require additional dependencies to be installed. In such cases, we can use the `BiocManager::install()` function from the Bioconductor project to install the necessary dependencies.

Here is an example code snippet to install the required packages:
you can uncomment the following code to install required packages
```{r}
# Install necessary packages
install.packages("gridExtra")
install.packages("jpeg")
install.packages("imager")
install.packages("magick")

# Install Bioconductor package (if not already installed)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install EBImage package from Bioconductor
BiocManager::install("EBImage")
install.packages("abind")

# Install torch, torchvision, and luz
install.packages("torch")
install.packages("torchvision")
install.packages("luz")
```

```{r}
# load packages
library(ggplot2)
library(gridExtra)
library(imager)
library(jpeg)
library(magick)
library(EBImage)
library(grid)
library(dplyr)
library(abind)
```

If using Mac OSX: imager needs [Xquartz](https://www.xquartz.org/)
- If using homebrew: `brew install --cask xquartz`

# Diagnosing Pneumonia from Chest X-Rays
## Parsing the data

Today, we are going to look at a series of chest X-rays from children (from [this paper](https://www.cell.com/cell/fulltext/S0092-8674(18)30154-5)) and see if we can accurately diagnose pneumonia from them.  

Chest X-ray images (anterior-posterior) were selected from retrospective cohorts of pediatric patients of one to five years old from Guangzhou Women and Children’s Medical Center, Guangzhou. All chest X-ray imaging was performed as part of patients’ routine clinical care. For the analysis of chest x-ray images, all chest radiographs were initially screened for quality control by removing all low quality or unreadable scans. The diagnoses for the images were then graded by two expert physicians before being cleared for inclusion. In order to account for any grading errors, the evaluation set was also checked by a third expert.

We can download, unzip, then inspect the format of this dataset as follows:

"https://drive.google.com/file/d/1-5JWvxlPayoXOuIhNPWkI0vGh4I8JfxF/view?usp=sharing"

you can **unzip** the file in a data directory, using the following code:
*uncomment it*

```{r}
# Specify the path to the zip file
zip_file <- "data/lab3_chest_xray_fix.zip"

# Specify the destination directory to extract the files
destination_dir <- "lab3_data"

# Extract the zip file
#unzip(zip_file, exdir = destination_dir)
```

As with every data analysis, the first thing we need to do is learn about our data. If we are diagnosing pneumonia, we should use the internet to better understand what pneumonia actually is and what types of pneumonia exist.

**0_** What is pneumonia and what is the point/benefit of being able to identify it from X-rays automatically?

Pneumonia is an infection that inflames the air sacs in one or both lungs, and fills them up with fluid or pus. It is caused by bacterial or fungal infection. It makes it difficult for the affected person to breathe and can cause a fever and cough with yellow, green or bloody mucus.

Automatically identifying pneumonia from an X-ray has several significant benefits as outlined below:

*Increased Diagnostic Accuracy:* AI models trained on large datasets can achieve high accuracy in identifying pneumonia, potentially reducing human error and improving diagnostic consistency.

*Efficiency and Speed:* Automated systems can quickly analyze X-rays, providing rapid results that can expedite the diagnosis process and facilitate faster treatment decisions, especially in urgent or emergency settings.

*Scalability:* AI can handle large volumes of X-rays efficiently, making it particularly useful in high-demand situations where there is a high patient load.

*Support for Under-Resourced Areas:* Automated diagnosis can also be beneficial in regions with limited access to radiologists, providing critical diagnostic support and improving healthcare accessibility.

*Standardization of Diagnosis:* Automated systems provide standardized interpretations of X-rays, reducing variability between different human radiologists and ensuring consistent diagnostic criteria.

*Radiologist Support:* AI can serve as a second opinion, supporting radiologists by highlighting potential areas of concern, thus reducing the cognitive load and allowing them to focus on complex cases.

*Resource Optimization:* By automating the initial screening process, healthcare resources can be optimized, allowing radiologists to focus on more critical tasks and improving overall workflow efficiency.

*Integration with Telemedicine:* Automated diagnostic tools can be integrated with telemedicine platforms, facilitating remote diagnosis and consultation, which is particularly beneficial for patients in remote or rural areas.

*Reduction in Diagnostic Disparities:* Automated systems can help reduce disparities in healthcare by providing consistent diagnostic capabilities across different regions and populations.

# Exploring dataset

In the provided code snippet, we are exploring the dataset directory structure for a chest x-ray image classification task. 

Make sure to put lab3_chest_xray directory in your R project directory.

```{r}
data_folder = "lab3_chest_xray"

files <- list.files(data_folder, full.names=TRUE, recursive=TRUE)
sort(sample(files, 20))
```

In the output above, you can see the filenames for all the chest X-rays. Look carefully at the filenames for the X-rays showing pneumonia (i.e., those in `{train,test}/PNEUMONIA`).

**1_** Do these filenames tell you anything about pneumonia? Why might this make predicting pneumonia more challenging?

The above file names indicate that these are chest X-rays from healthy individuals (normal) and patients with pneumonia (virus/bacteria). There are 14 training sets and 6 test sets, which means that the entire dataset was split into 70% for training and the rest 30% for testing. This split has also been done in terms of normal patients, patients with viral pneumonia and patients with bacterial pneumonia.

However, there appears to be more viral or bacterial test sets compared to the singular normal test set even though there are seven normal training sets. Since there are typically more normal X-rays than pneumonia cases, models trained in this data may become biased toward the majority class (normal), affecting their ability to detect pneumonia accurately.

Moreover, even though there are enough viral/bacterial pneumonia X-ray data in both training and test sets, this may still pose as a challenge. Pneumonia can manifest differently in X-rays, making it challenging to create a one-size-fits-all model. There may also be overlapping diseases which manifest through symptoms of pneumonia. If pneumonia is not properly distinguished from these diseases in the model, it may create more harm for the patient than good. 

The dataset folder is assumed to contain three subfolders: train, test, and validate. Each of these subfolders further contains two folders representing the two classes or labels in our classification task: "normal" and "pneumonia".

The code provides a way to explore the dataset directory structure and obtain information about the number of images and their distribution across different classes. This information can be helpful in understanding the dataset and planning the subsequent steps of the image classification task.

```{r}
# Exploring dataset
base_dir <- "lab3_chest_xray"

train_pneumonia_dir <- file.path(base_dir, "train", "PNEUMONIA")
train_normal_dir <- file.path(base_dir, "train", "NORMAL")

test_pneumonia_dir <- file.path(base_dir, "test", "PNEUMONIA")
test_normal_dir <- file.path(base_dir, "test", "NORMAL")

val_normal_dir <- file.path(base_dir, "validate", "NORMAL")
val_pneumonia_dir <- file.path(base_dir, "validate", "PNEUMONIA")

train_pn <- list.files(train_pneumonia_dir, full.names = TRUE)
train_normal <- list.files(train_normal_dir, full.names = TRUE)

test_normal <- list.files(test_normal_dir, full.names = TRUE)
test_pn <- list.files(test_pneumonia_dir, full.names = TRUE)

val_pn <- list.files(val_pneumonia_dir, full.names = TRUE)
val_normal <- list.files(val_normal_dir, full.names = TRUE)

cat("Total Images:", length(c(train_pn, train_normal, test_normal, test_pn, val_pn, val_normal)), "\n")
cat("Total Pneumonia Images:", length(c(train_pn, test_pn, val_pn)), "\n")
cat("Total Normal Images:", length(c(train_normal, test_normal, val_normal)), "\n")

```
# Creating training datasets

The provided code segment focuses on creating datasets for training, testing, and validation, as well as assigning labels to the corresponding datasets. Additionally, the code shuffles the data to introduce randomness in the order of the samples. Here's a breakdown of the code:

`train_dataset`: Combines the lists `train_pn` and `train_normal` to create a single dataset for training.

`train_labels`: Creates a vector of labels for the training dataset by repeating "pneumonia" for the length of `train_pn` and "normal" for the length of `train_normal`.

`shuffled_train_dataset`, `shuffled_train_labels`: Extracts the shuffled training dataset and labels from the shuffled `train_data` data frame.

By creating datasets and assigning labels, this code prepares the data for subsequent steps, such as model training and evaluation. The shuffling of the data ensures that the samples are presented in a random order during training, which can help prevent any biases or patterns that may exist in the original dataset.
```{r}
train_dataset <- c(train_pn, train_normal)
train_labels <- c(rep("pneumonia", length(train_pn)), rep("normal", length(train_normal)))

test_dataset <- c(test_pn, test_normal)
test_labels <- c(rep("pneumonia", length(test_pn)), rep("normal", length(test_normal)))

val_dataset <- c(val_pn, val_normal)
val_labels <- c(rep("pneumonia", length(val_pn)), rep("normal", length(val_normal)))

# Create a data frame with the dataset and labels
train_data <- data.frame(dataset = train_dataset, label = train_labels)
test_data <- data.frame(dataset = test_dataset, label = test_labels)
val_data <- data.frame(dataset = val_dataset, label = val_labels)

# Shuffle the data frame
train_data <- train_data[sample(nrow(train_data)), ]
test_data <- test_data[sample(nrow(test_data)), ]
val_data <- val_data[sample(nrow(val_data)), ]

# Extract the shuffled dataset and labels
shuffled_train_dataset <- train_data$dataset
shuffled_train_labels <- train_data$label

shuffled_test_dataset <- test_data$dataset
shuffled_test_labels <- test_data$label

shuffled_val_dataset <- val_data$dataset
shuffled_val_labels <- val_data$label
```

```{r}
#showing a file name from test set
cat("file name: ", shuffled_train_dataset[5], "\nlabel: ", shuffled_train_labels[5])
```

# Data Visualization

Let’s inspect a couple of these files as it is always worth looking directly at data.

```{r}
# Create a list to store the ggplot objects
plots <- list()

# Iterate through the images and labels
for (i in 1:4) {
  
  image <- readImage(shuffled_train_dataset[i])
  
  # Create a ggplot object for the image with the corresponding label
  plot <- ggplot() +
    theme_void() +
    annotation_custom(
      rasterGrob(image, interpolate = TRUE),
      xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
    )
  
  # Add the ggplot object to the list
  plots[[i]] <- plot
}

# Arrange the plots in a 2x2 grid
grid.arrange(grobs = plots, nrow = 2, ncol = 2)
```

**2_** From looking at only 3 images do you see any attribute of the images that we may have to normalise before training a model?

From looking at the images, we may have to normalize the following attributes to get proper results from a model:

*Image Size and Resolution:* It appears that the images are not of the same dimension and there are also differences in their resolutions. Normalizing these images by maintaining consistent size and pixel resolution is imperative to properly train a model.

*Brightness or Contrast:* The brightness of one of the images also appears to be higher than the rest, which needs to be adjusted to minimize variations between the images.

*Patient Positioning:* The patients in the images also appear to be positioned slightly differently which may not create any issues for doctors in reading the X-rays, but might pose a challenge for an ML model.

# Data Pre-processing

The provided code segment presents a function called `process_images` that performs several preprocessing steps on the images. Here's an explanation of each preprocessing step and its purpose:

1. Loading the "imager" library: This line imports the "imager" library, which provides functions for image processing.

2. Setting the desired image size: The variable `img_size` is initialized to 224, indicating the desired size (both width and height) of the processed images. This step ensures that all images are resized to a consistent size.

3. Initializing an empty list for processed images: The variable `X` is initialized as an empty list that will store the processed images.

4. Looping through each image path: The function iterates over each image path in the `shuffled_dataset` input, which represents the paths to the shuffled images.

5. Loading the image: The `imager::load.image()` function is used to read and load the image from its file path.

6. Normalizing the image: The loaded image is divided by 255 to normalize its pixel values. This step scales the pixel values between 0 and 1, which is a common practice in image processing and deep learning.

7. Resizing the image: The `resize()` function from the "imager" library is employed to resize the normalized image to the desired `img_size`. Resizing the images to a consistent size is important for ensuring compatibility with the subsequent steps of the deep learning pipeline.

8. Appending the processed image to the list: The processed image, stored in the variable `img_resized`, is added to the `X` list using the `c()` function and the `list()` function. The resulting `X` list will contain all the processed images.

9. Returning the processed images: Finally, the `X` list, which now holds the processed images, is returned as the output of the `process_images` function.

These preprocessing steps are commonly performed in image classification tasks to prepare the images for training a deep learning model. Normalizing the pixel values and resizing the images ensure that they are in a consistent format and range, which facilitates the learning process of the model. Additionally, resizing the images to a fixed size allows for efficient batch processing and ensures that all images have the same dimensions, enabling them to be fed into the model's input layer.

```{r}
process_images <- function(shuffled_dataset) {
  
  img_size <- 224  # Desired image size
  
  # Initialize an empty list to store processed images
  X <- list()
  
  # Loop through each image path in shuffled_train_dataset
  for (image_path in shuffled_dataset) {
    # Read the image
    img <- imager::load.image(image_path)
    
    # Normalize the image
    img_normalized <- img / 255
    
    # Resize the image
    img_resized <- resize(img_normalized, img_size, img_size)
    
    # Append the processed image to the list
    X <- c(X, list(img_resized))
  }
  
  return(X)
}
```

In this section, by using the process_images function we will do preprocessing on the training, testing, and validation images.

Then we will encode the labels:
The ifelse() function is utilized to encode the labels. If a label in shuffled_train_labels, shuffled_test_labels, or shuffled_val_labels is "normal," it is assigned a value of 1. Otherwise, if the label is "pneumonia," it is assigned a value of 2. This encoding scheme allows for easier handling of the labels in subsequent steps.

Finally, We Convert labels to integer type:
The labels are converted to the integer data type using the as.integer() function. This ensures that the labels are represented as integers, which is the expected format for the target tensor when using the nn_cross_entropy_loss function.

It is important to note that when using the nn_cross_entropy_loss function in R, the target tensor is expected to have a "long" data type, which is equivalent to the integer type in R. Hence, the labels need to be converted to integers.

Furthermore, when working with the Torch package in R, it is essential to ensure that labels start from 1 instead of 0. In binary classification problems, the labels should be 1 and 2, representing the two classes. This adjustment is necessary to avoid errors when using the labels as indices for the output tensor.


```{r}
train_X <- process_images(shuffled_train_dataset)
test_X <- process_images(shuffled_test_dataset)
val_X <- process_images(shuffled_val_dataset)

train_y <- ifelse(shuffled_train_labels == "normal", 1, 2)
test_y <- ifelse(shuffled_test_labels == "normal", 1, 2)
val_y <- ifelse(shuffled_val_labels == "normal", 1, 2)

train_y <- as.integer(train_y)
test_y <- as.integer(test_y)
val_y <- as.integer(val_y)
```

Now let's have an other look at images after doing pre processing.

```{r}
# Create a list to store the ggplot objects
plots <- list()

# Iterate through the images and labels
for (i in 1:4) {
  if (train_y[i] == 0) {
    label <- "Normal"
  } else {
    label <- "Pneumonia"
  }
  
  # Create a ggplot object for the image with the corresponding label
  plot <- ggplot() +
    theme_void() +
    ggtitle(label) +
    annotation_custom(
      rasterGrob(train_X[[i]], interpolate = TRUE),
      xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
    )
  
  # Add the ggplot object to the list
  plots[[i]] <- plot
}

# Arrange the plots in a 2x2 grid
grid.arrange(grobs = plots, nrow = 2, ncol = 2)
```

Let's count and display the number of images in each dataset to examine the distribution of labels in the data. This will help us understand the ratio of "normal" and "pneumonia" labels in the dataset.

```{r}
# Combine train, test, and val vectors into a single data frame
df <- data.frame(
  Data = rep(c("Train", "Test", "Val"), times = c(length(train_y), length(test_y), length(val_y))),
  Value = c(train_y, test_y, val_y)
)

# Create a single bar plot with facets
fig <- ggplot(df, aes(x = Value)) +
  geom_bar() +
  ylim(0, 510) +
  facet_wrap(~Data, ncol = 3)

# Arrange the plot
grid.arrange(fig, nrow = 1)

```

As you can see, the provided dataset is completely balanced in all sets.

**3_** if the dataset was not balanced, what kind of techniques could be useful?

If the dataset was not balanced, we could use a number of techniques to balance out the data and make a more effective model. These techniques would depend on the type of data we have and outcome after implementation. Some of the techniques that could work for our data are briefly described below:

*Oversampling and/or Undersampling:* We could increase the minority class data by oversampling and decrease the majority class data by undersampling to balance out the dataset.

*Weight Distribution:* We could change the weights associated with each class to compensate for the imbalance.

*Data Augmentation:* To account for the minority class, we could create additional data points by modifying the existing data.

*Ensemble Techniques:* We could also utilize multiple models like bagging, boosting, or stacking and combine the predictions to help in the overall model performance.


# Training

In the next step, we need to reshape the dataset to ensure it is in the appropriate format for feeding into deep learning models. Reshaping involves modifying the structure and dimensions of the data to match the expected input shape of the model. 

```{r}
train_X <- array(data = unlist(train_X), dim = c(1000, 224, 224, 1))
test_X <- array(data = unlist(test_X), dim = c(200, 224, 224, 1))
val_X <- array(data = unlist(val_X), dim = c(16, 224, 224, 1))
```

```{r}
print(dim(train_X))
print(length(train_y))
print(dim(test_X))
print(length(test_y))
print(dim(val_X))
print(length(val_y))
```
The last dimension of an image represents the color channels, typically RGB (Red, Green, Blue) in color images or grayscale in black and white images. In our case, since we are working with black and white images, there is only one channel, hence the value 1.

However, the deep learning framework expects the input tensor to have a specific shape, with the color channels as the second dimension. The desired shape is (batch_size x channels x height x width), but our current data has the shape (batch_size x height x width x channels).

To rearrange the dimensions of our data to match the expected shape, we use the aperm function in R. The aperm function allows us to permute the dimensions of an array. In this case, we are permuting the dimensions of train_X to change the order of the dimensions, so that the channels dimension becomes the second dimension.

```{r}
train_X <- aperm(train_X, c(1,4,2,3))
test_X <- aperm(test_X, c(1,4,2,3))
val_X <- aperm(val_X,c(1,4,2,3))

dim(train_X)
```

In order to train a Convolutional Neural Network (CNN), we will utilize the torch package in R. Torch is a powerful deep learning library that provides a wide range of functions and tools for building and training neural networks.

CNNs are particularly effective for image-related tasks due to their ability to capture local patterns and spatial relationships within the data. Torch provides a high-level interface to define and train CNN models in R.

By using torch, we can leverage its extensive collection of pre-built layers, loss functions, and optimization algorithms to construct our CNN architecture. We can define the network structure, specifying the number and size of convolutional layers, pooling layers, and fully connected layers.

```{r}
# Load the torch package
library(torch)
library(torchvision)
library(luz)
```

In this code, we are defining a custom dataset class called "ImageDataset" to encapsulate our training, testing, and validation data. The purpose of the dataset class is to provide a structured representation of our data that can be easily consumed by deep learning models.

The "ImageDataset" class has three main functions:
1. "initialize": This function is called when creating an instance of the dataset class. It takes the input data (X) and labels (y) as arguments and stores them as tensors.
2. ".getitem": This function is responsible for retrieving a single sample and its corresponding label from the dataset. Given an index (i), it returns the i-th sample and label as tensors.
3. ".length": This function returns the total number of samples in the dataset.

After defining the dataset class, we create instances of it for our training, testing, and validation data: "train_dataset", "test_dataset", and "val_dataset". We pass the respective input data and labels to each dataset instance.

Next, we create dataloader objects for each dataset. A dataloader is an abstraction that allows us to efficiently load and iterate over the data in batches during the training process. The batch size (16 in this case) determines the number of samples that will be processed together in each iteration. It helps in optimizing memory usage and can speed up the training process by leveraging parallel computation.

Finally, the code visualizes the size of the first batch by calling "batch[[1]]$size()". This can be useful for understanding the dimensions of the data and ensuring that the input shapes are consistent with the network architecture.

```{r}
# Define a custom dataset class
ImageDataset <- dataset(
  name = "ImageDataset",
  initialize = function(X, y) {
    # Store the data as tensors
    self$data <- torch_tensor(X)
    self$labels <- torch_tensor(y)
  },
  .getitem = function(i) {
    # Return a single sample and label
    x <- self$data[i,,,]
    y <- self$labels[i]
    list(x = x, y = y)
  },
  .length = function() {
    # Return the number of samples
    dim(self$data)[1]
  }
)

# Create a dataset object from your data
train_dataset <- ImageDataset(train_X, train_y)
test_dataset <- ImageDataset(test_X, test_y)
val_dataset <- ImageDataset(val_X, val_y)

# Create a dataloader object from your dataset
train_dataloader <- dataloader(train_dataset, batch_size = 16)
test_dataloader <- dataloader(test_dataset, batch_size = 16)
val_dataloader <- dataloader(val_dataset, batch_size = 16)

# Iterate over batches of data
batch = train_dataloader$.iter()$.next()

# Visualize the first batch size
batch[[1]]$size()
```
## Create the CNN model

The input image has one channel and a size of 224 x 224 pixels. The first convolutional layer has 32 filters with a kernel size of 3 x 3 and a stride of 1. The output of this layer has a size of 32 x 222 x 222. The second convolutional layer has 64 filters with the same kernel size and stride. The output of this layer has a size of 64 x 220 x 220. The max pooling layer has a kernel size of 2 x 2 and reduces the spatial dimensions by half. The output of this layer has a size of 64 x 110 x 110. The dropout layer randomly sets some elements to zero with a probability of 0.25. The flatten layer reshapes the output into a vector with a length of 774400. The first fully connected layer has 128 neurons and applies a ReLU activation function. The second dropout layer randomly sets some elements to zero with a probability of 0.5. The second fully connected layer has 2 neurons and produces the final output for the classification task.

Input Image: 1 channel, 224 x 224

| Layer Type    | Output Size    | Parameters      |
| ------------- | -------------- | --------------- |
| Conv2D        | 32 x 222 x 222 | 32 x 3 x 3      |
|               |                | (weights)       |
| Conv2D        | 64 x 220 x 220 | 64 x 3 x 3      |
|               |                | (weights)       |
| MaxPooling2D  | 64 x 110 x 110 | 2 x 2           |
|               |                | (kernel size)   |
| Dropout       | 64 x 110 x 110 | 0.25            |
|               |                | (dropout rate)  |
| Flatten       | 774400         | -               |
| FullyConnected| 128            | -               |
| (ReLU)        |                |                 |
| Dropout       | 128            | 0.5             |
|               |                | (dropout rate)  |
| FullyConnected| 2              | -               |



```{r}
net <- nn_module(
  "Net",
  
  initialize = function() {
    self$conv1 <- nn_conv2d(1, 32, 3, 1)
    self$conv2 <- nn_conv2d(32, 64, 3, 1)
    self$dropout1 <- nn_dropout2d(0.25)
    self$dropout2 <- nn_dropout2d(0.5)
    self$fc1 <- nn_linear(774400, 128)  # Adjust the input size based on your image dimensions
    self$fc2 <- nn_linear(128, 2)             # Change the output size to match your classification task
  },
  
  forward = function(x) {
    x %>%                                        # N * 1 * 224 * 224
      self$conv1() %>%                           # N * 32 * 222 * 222
      nnf_relu() %>% 
      self$conv2() %>%                           # N * 64 * 220 * 220
      nnf_relu() %>% 
      nnf_max_pool2d(2) %>%                      # N * 64 * 110 * 110
      self$dropout1() %>% 
      torch_flatten(start_dim = 2) %>%           # N * 64 * 110 * 110 --> N * 774400
      self$fc1() %>%                             # N * 128
      nnf_relu() %>% 
      self$dropout2() %>% 
      self$fc2()                                 # N * 2 (change the output size to match your classification task)
  }
)

```

## Train model

```{r warning=FALSE}
# Set the number of epochs
num_epochs <- 3

train_loss <- numeric(num_epochs)
train_acc <- numeric(num_epochs)
test_loss <- numeric(num_epochs)
test_acc <- numeric(num_epochs)

# Loop through the epochs
for (epoch in 1:num_epochs) {
  # Perform training and validation for each epoch
  fitted <- net %>%
    setup(
      loss = nn_cross_entropy_loss(),
      optimizer = optim_adam,
      metrics = list(
        luz_metric_accuracy()
      )
    ) %>%
    fit(train_dataloader, epochs = 1, valid_data = test_dataloader)
  
  # Print the metrics for the current epoch
  cat("Epoch ", epoch, "/", num_epochs, "\n")
  cat("Train metrics: Loss: ", fitted$records$metrics$train[[1]]$loss, " - Acc: ", fitted$records$metrics$train[[1]]$acc, "\n")
  cat("Valid metrics: Loss: ", fitted$records$metrics$valid[[1]]$loss, " - Acc: ", fitted$records$metrics$valid[[1]]$acc, "\n")
  cat("\n")
  
  # Store the loss and accuracy values
  train_loss[epoch] <- fitted$records$metrics$train[[1]]$loss
  train_acc[epoch] <- fitted$records$metrics$train[[1]]$acc
  test_loss[epoch] <- fitted$records$metrics$train[[1]]$loss
  test_acc[epoch] <- fitted$records$metrics$valid[[1]]$acc
}

```
## Plot learning curves

```{r}
# Plot the train and test loss
loss_df <- data.frame(
  Epoch = 1:num_epochs,
  Train_Loss = train_loss,
  Test_Loss = test_loss
)
loss_plot <- ggplot(data = loss_df) +
  geom_line(aes(x = Epoch, y = Train_Loss, color = "Train Loss")) +
  geom_line(aes(x = Epoch, y = Test_Loss, color = "Test Loss")) +
  labs(x = "Epoch", y = "Loss") +
  scale_color_manual(values = c("Train Loss" = "blue", "Test Loss" = "red")) +
  theme_minimal()

# Plot the train and test accuracy
acc_df <- data.frame(
  Epoch = 1:num_epochs,
  Train_Accuracy = train_acc,
  Test_Accuracy = test_acc
)
acc_plot <- ggplot(data = acc_df) +
  geom_line(aes(x = Epoch, y = Train_Accuracy, color = "Train Accuracy")) +
  geom_line(aes(x = Epoch, y = Test_Accuracy, color = "Test Accuracy")) +
  labs(x = "Epoch", y = "Accuracy") +
  scale_color_manual(values = c("Train Accuracy" = "blue", "Test Accuracy" = "red")) +
  theme_minimal()

# Print the plots
print(loss_plot)
print(acc_plot)
```

**4_** Based on the training and test accuracy, is this model actually managing to classify X-rays into pneumonia vs normal? What do you think contributes to this? why? (not required to answer)

It seems that the model is not training effectively the model is struggling to learn from the data and is not generalizing well to unseen examples. It suggests that the model may be too complex or the training process needs adjustments.

**5_** what is your suggestions to solve this problem? How could we improve this model? (not required to answer)

Since the model is too complex and not performing as well as it was intended, we can take utilize a number of techniques to solve the problem and improve this model.

*Simplify the Model:* Since the current model is too complex and hence leading to overfitting, we can try to simplify the model architecture by implementing fewer layers, fewer parameters, and applying regularization techniques like dropout, L1/L2 regularization, or batch normalization.

*Hyperparameter Tuning:* Another way we could improve the performance of the model is through playing around with the different hyperparameters. We could adjust the learning rates and batch size to find the right balance between stability and convergence.

*Data Augmentation:* We can also generate additional training samples by applying transformations (rotation, scaling, flipping) to existing X-rays. Augmented data usually helps the model to generalize better.

*Cross-Validation:* Another options is to use k-fold cross-validation to assess model stability and generalization. This will validate the model on different subsets of the data.

In conclusion, iterative experimentation with different techniques and fine-tuning with paramaters until we achieve satisfactory performance from the model is the best approach when it comes to complex models.

# Data Augmentation
Data augmentation is one of the solutions to consider when facing training difficulties. Data augmentation involves applying various transformations or modifications to the existing training data, effectively expanding the dataset and introducing additional variations. This technique can help improve model performance and generalization by providing the model with more diverse examples to learn from.

By applying data augmentation, we can create new training samples with slight modifications, such as random rotations, translations, flips, zooms, or changes in brightness and contrast. These modifications can mimic real-world variations and increase the model's ability to handle different scenarios. Data augmentation increased dataset size, improved generalization, and reduced overfitting.

Below is an example that demonstrates how we can augment the data.

```{r}
img <- readImage(shuffled_train_dataset[5])

T_img <- torch_squeeze(torch_tensor(img)) %>% 
        # Randomly change the brightness, contrast and saturation of an image
        transform_color_jitter() %>%
        # Horizontally flip an image randomly with a given probability
        transform_random_horizontal_flip() %>%
        # Vertically flip an image randomly with a given probability
        transform_random_vertical_flip(p = 0.5)

image(as.array(T_img))
```

We can also add the data transformations to our dataloader:

```{r}
# Define a custom dataset class with transformations
ImageDataset_augment <- dataset(
  name = "ImageDataset",
  initialize = function(X, y, transform = NULL) {
    self$transform <- transform
    self$data <- X
    self$labels <- y
  },
  .getitem = function(i) {
    # Return a single sample and label
    x <- self$data[i,,,]
    x <- self$transform(x)
    y <- self$labels[i]
    
    list(x = x, y = y)
  },
  .length = function() {
    dim(self$data)[1]
  }
)

# Define the transformations for training data
train_transforms <- function(img) {
  img <- torch_squeeze(torch_tensor(img)) %>%
    transform_color_jitter() %>%
    transform_random_horizontal_flip() %>%
    transform_random_vertical_flip(p = 0.5) %>%
    torch_unsqueeze(dim = 1)
  
  return(img)
}

# Apply the transformations to your training dataset
train_dataset <- ImageDataset_augment(train_X, train_y, transform = train_transforms)
test_dataset <- ImageDataset(test_X, test_y)
val_dataset <- ImageDataset(val_X, val_y)

# Create a dataloader for training
train_dataloader <- dataloader(train_dataset, batch_size = 16)
test_dataloader <- dataloader(test_dataset, batch_size = 16)
val_dataloader <- dataloader(val_dataset, batch_size = 16)

# Iterate over batches of data
batch = train_dataloader$.iter()$.next()

# Visualize the first batch size
batch[[1]]$size()
```

**6_** What are the potential drawbacks or disadvantages of data augmentation?

Data augmentation is a technique to expand the dataset to enhance model performance, but it also has potential drawbacks when applied to X-ray data for pneumonia prediction:

*Introduction of Artificial Patterns:* Data augmentation introduces variations by applying transformations (e.g., flipping, rotation, brightness changes). However, these variations may not always reflect real-world anatomical or pathological differences. These patterns could mislead the model during training.

*Risk of Overfitting:* Augmented data increases the effective size of the training dataset. However, if the augmentation is too aggressive, the model may overfit to the augmented samples. Balancing the augmentation with regularization techniques is crucial in this case.

*Preserving Anatomical Integrity:* X-rays capture specific anatomical structures (e.g., lung shape, ribcage). Augmentation should not distort these critical features.

*Noise Amplification:* Augmentation can sometimes introduce noise (e.g., random brightness changes). If the noise is excessive, it might hinder the model’s ability to learn relevant features.

*Interpretability Challenges:* Augmented images may be harder to interpret by clinicians.
Unusual augmentations could confuse human experts during diagnosis and therefore the entire purpose of having dual confirmation of a diagnosis would not be achieved.


# Mobile net

Transfer learning is another technique in machine learning where knowledge gained from solving one problem is applied to a different but related problem. It involves leveraging pre-trained models that have been trained on large-scale datasets and have learned general features. By utilizing transfer learning, models can benefit from the knowledge and representations learned from these pre-trained models.

Torchvison provides versions of all the integrated architectures that have already been trained on the ImageNet dataset.

**7_** What is ImageNet aka “ImageNet Large Scale Visual Recognition Challenge 2012”? How many images and classes does it involve? Why might this help us?

The ImageNet Large Scale Visual Recognition Challenge 2012 (ILSVRC2012) was a pivotal competition in the field of computer vision. The goal of this competition was to estimate the content of images for the purpose of retrieval and automatic annotation using a subset of the large hand-labeled ImageNet dataset and identify the main objects present in the images.

*Dataset:* The competition used only a subset of the ImageNet dataset which contains over 10 million labeled images across 10,000+ object categories. This dataset or subset of the dataset served as training data for the challenge.

*Aim of the Competition:* Participants aimed to estimate the content of the images for retrieval and automatic annotation, where test images were presented without initial annotations (no labels or segmentation), and algorithms had to produce labels indicating the objects present in the images. In addition to classification, there was a detection task where algorithms had to specify the location of objects.

*Significance of the Competition:* Most participants used AlexNet, a deep convolutional neural network (CNN) for their models and its performance revolutionized the field. The challenge spurred research in large-scale image classification and localization, leading to advancements in deep learning and CNN architectures. It also provided a benchmark for evaluating progress in object detection and classification across diverse objects.

Considering that this dataset and the ILSVRC2012 competition played a crucial role in shaping the machine learning field we know today, it helps us in understanding how the development of this field and our role in continuing the research in the future.

MobileNet is a pre-trained model that can be effectively utilized in transfer learning scenarios. Do some research about this model.

**8_** Why do you think using this architecture in this practical assignment can help to improve the results? 
Hint: See MobileNet publication

MobileNet is an efficient convolutional neural network (CNN) architecture designed specifically for mobile and embedded vision applications. It is a computer-vision model open-sourced by Google and designed for training classifiers.

Let’s delve into its key features and understand why this architecture can help us to improve results:

*Depth-wise Separable Convolution:* MobileNet employs a depth-wise separable convolution approach. It consists of two layers - Depth-wise convolution, which applies a single filter per input channel and reduces computation, and Point-wise convolution, which combines depth-wise features across channels. This separation significantly reduces the number of parameters and computations compared to standard convolutions.

*Efficiency and Lightweight Design:* MobileNet’s streamlined architecture builds lightweight deep neural networks. It balances latency and accuracy using simple global hyper-parameters. Developers can choose the right-sized model based on their application constraints.

*Advantages for Transfer Learning:* When applied in transfer learning scenarios, MobileNet’s lightweight design allows it to adapt well to new tasks. It retains competitive accuracy while being mindful of resource limitations. Fine-tuning MobileNet on specific datasets can yield impressive results.

*Practical Assignment Benefits:* Using MobileNet in your practical assignment can lead to:
1. Faster training and inference: This model can significantly reduce computational overhead considering the huge number of parameters associated with medical data. 
2. Adaptability: Since this model is adaptable, it can utilize transfer learning to make it adaptable to specific tasks as well as create compatibility with mobile or edge devices.
3. Good accuracy: Early detection is essential for effective treatment and improved patient outcomes when it comes to pneumonia. This model is especially helpful when fine-tuning relevant data to achieve desirable and accurate results.
4. Robustness and Variability: Chest X-ray images exhibit variability due to acquisition and processing differences. Since MobileNet is trained on large and high-quality datasets, it can handle the variability that comes with Chest X-ray images.

In summary, MobileNet’s efficiency, accuracy, and adaptability make it a valuable tool for pneumonia detection in chest X-ray images and hence can help our results in this practical.

**9_** How many parameters does this network have? How does this compare to better performing networks available in torch?

The MobileNet architecture has varying parameter counts based on its configuration:

Standard MobileNet: The base MobileNet model contains approximately 4.2 million parameters. It strikes a balance between efficiency and accuracy.

Smaller Versions: Smaller variants of MobileNet exist, with around 1.32 million parameters. These versions maintain efficiency while reducing the model size.

*Comparison to Other Networks:*

MobileNet’s parameter count is significantly lower than heavyweight networks like ResNet or VGG. While MobileNet sacrifices some accuracy, it excels in resource-constrained scenarios. MobileNet’s parameter efficiency makes it a valuable choice for mobile vision applications, especially when compared to larger networks. 

Having said that, it's important to choose an architecture best suited for a specific task. While ResNet is excellent in accuracy but heavier in terms of parameters and TabNet is ideal for tabular data analysis and interpretability, MobileNet is more efficient and suitable for mobile applications. Moreover, we can fine-tune MobileNet on your specific dataset to improve its performance as well as leverage pre-trained weights to allow for transfer learning.


In the following you can see an example of how we can load pre-trained models in our codes.

```{r}
# Load the pre-trained MobileNet model

mobilenet<-model_mobilenet_v2(pretrained = TRUE)

# Modify the last fully connected layer to match your classification task

#in_features <- mobilenet$classifier$in_features
mobilenet$classifier <- nn_linear(224*224*3, 2)   # Adjust the output size based on your classification task
mobilenet
```
**10_** Using the provided materials in this practical, train a different network architecture. Does this perform better? (not required to answer)

# Useful links
https://medium.com/@kemalgunay/getting-started-with-image-preprocessing-in-r-52c7d153b381
https://cran.r-project.org/web/packages/magick/vignettes/intro.html
https://www.datanovia.com/en/blog/easy-image-processing-in-r-using-the-magick-package/
https://dahtah.github.io/imager/imager.html
https://rdrr.io/github/mlverse/torchvision/api/
https://github.com/brandonyph/Torch-for-R-CNN-Example
