---
title: "Practical 4"
author: "Rifat Fariha B00937648"
date: "2024-06-15"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
There are 3-4 packages you will need to install for today's practical: `install.packages(c("xgboost", "eegkit", "forecast", "tseries", "caret"))` apart from that everything else should already be available on your system. 

If you are using a newer Mac you may have to also install [quartz](https://www.xquartz.org/) to have everything work (do this if you see errors about `X11` during install/execution).

I will endeavour to use explicit imports to make it clear where functions are coming from (functions without `library_name::` are part of base R or a function we've defined in this notebook).

```{r libraries, echo=FALSE}
# Using the same library we used earlier in the course for tabular data because we know it works!
library(xgboost)

# EEG manipulation library in R (although very limited compared to signal processing libraries available in other languages, matlab might actually still be a leader in this specific area)
library(eegkit)

# some time series functions (that we only skim the depths of)
library(forecast)
library(tseries)
library(caret)

# just tidyverse libraries that should already be installed
library(dplyr)
library(reshape2)
library(purrr)
library(ggplot2)
```

## EEG Eye Detection Data

One of the most common types of medical sensor data (and one that we talked about during the lecture) are Electroencephalograms (EEGs).  
These measure mesoscale electrical signals (measured in microvolts) within the brain, which are indicative of a region of neuronal activity.
Typically, EEGs involve an array of sensors (aka channels) placed on the scalp with a high degree of covariance between sensors.

As EEG data can be very large and unwieldy, we are going to use a relatively small/simple dataset today from [this paper](http://ehrai.com/su/pdf/aihls2013.pdf).

This dataset is a 117 second continuous EEG measurement collected from a single person with a device called a "Emotiv EEG Neuroheadset".
In combination with the EEG data collection, a camera was used to record whether person being recorded had their eyes open or closed. 
This was eye status was then manually annotated onto the EEG data with `1` indicated the eyes being closed and `0` the eyes being open.
Measures microvoltages are listed in chronological order with the first measured value at the top of the dataframe.

Let's parse the data directly from the `h2o` library's (which we aren't actually using directly) test data S3 bucket:

```{r parse_data}
eeg_url <- "https://h2o-public-test-data.s3.amazonaws.com/smalldata/eeg/eeg_eyestate_splits.csv"
eeg_data <- read.csv(eeg_url)

# add timestamp
Fs <- 117 / nrow(eeg_data)
eeg_data <- transform(eeg_data, ds = seq(0, 116.99999, by = Fs), eyeDetection = as.factor(eyeDetection))
print(table(eeg_data$eyeDetection))

# split dataset into train, validate, test
eeg_train <- subset(eeg_data, split == 'train', select = -split)
print(table(eeg_train$eyeDetection))

eeg_validate <- subset(eeg_data, split == 'valid', select = -split)
eeg_test <- subset(eeg_data, split == 'test', select = -split)
```

**0** Knowing the `eeg_data` contains 117 seconds of data, inspect the `eeg_data` dataframe and the code above to and determine how many samples per second were taken?

```{r}
# The duration of the recording in seconds
recording_duration <- 117

# Calculate the sampling frequency
sampling_frequency <- n_samples / recording_duration

# Print the sampling frequency
sampling_frequency
```
We can see that approximately 128 samples per second was taken.

**1** How many EEG electrodes/sensors were used?

To determine the number of EEG electrodes/sensors used in the dataset, we should first inspect the columns of the eeg_data dataframe. Given that the eeg_data dataframe also contains columns like split, eyeDetection, and the added ds (timestamp) column, we need to exclude these from our count of sensor columns. Then, we can just print the number of sensor columns.

```{r}
# Display the column names
colnames(eeg_data)

# Exclude the columns that are not sensors
sensor_columns <- setdiff(colnames(eeg_data), c("split", "eyeDetection", "ds"))

# Count the number of sensor columns
n_sensors <- length(sensor_columns)

# Print the number of sensors
n_sensors
```
As we can see, 14 EEG electrodes/sensors were used. 

### Exploratory Data Analysis

Now that we have the dataset and some basic parameters let's begin with the ever important/relevant exploratory data analysis.

First we should check there is no missing data!
```{r check_na}
sum(is.na(eeg_data))
```

Great, now we can start generating some plots to look at this data within the time-domain.

First we use `reshape2::melt()` to transform the `eeg_data` dataset from a wide format to a long format expected by `ggplot2`.

Specifically, this converts from "wide" where each electrode has its own column, to a "long" format, where each observation has its own row. 
This format is often more convenient for data analysis and visualization, especially when dealing with repeated measurements or time-series data.

We then use `ggplot2` to create a line plot of electrode intensities per sampling time, with the lines coloured by electrode, and the eye status annotated using dark grey blocks.

```{r plot_data}
melt <- reshape2::melt(eeg_data %>% dplyr::select(-split), id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")


ggplot2::ggplot(melt, ggplot2::aes(x=ds, y=microvolts, color=Electrode)) + 
  ggplot2::geom_line() + 
  ggplot2::ylim(3500,5000) + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(melt, eyeDetection==1), alpha=0.005)
```

**2** Do you see any obvious patterns between eyes being open (dark grey blocks in the plot) and the EEG intensities?

Considering the dark grey blocks to indicate periods when the eyes are open, there doesn’t appear to be a consistent or obvious pattern between eye openness and EEG intensities across all electrodes.

However, when the eyes seem to be open for longer periods of time, it seems EEG signal fluctuations subside and become more consistent. This could be because when eyes are open, the brain is actively processing visual information from the environment. Visual input stimulates various brain regions, including the visual cortex. This activation can influence other brain areas, leading to more synchronized neural activity. As a result, EEG signals may appear less noisy during open-eye periods.

**3** Similarly, based on the distribution of eye open/close state over time to anticipate any temporal correlation between these states?

To anticipate any temporal correlation between eye open/close states based on the distribution of these states over time, we can calculate and plot the proportion of time spent in each state over different time intervals. This can help in identifying any patterns or cycles in the data.

```{r}
# Calculate the proportion of time spent in each state over different time intervals
time_intervals <- seq(0, max(eeg_data$ds), by = 1)  # 1 second intervals
state_distribution <- eeg_data %>%
  dplyr::mutate(time_interval = cut(ds, breaks = time_intervals)) %>%
  dplyr::group_by(time_interval) %>%
  dplyr::summarize(
    proportion_closed = mean(eyeDetection == 1),
    proportion_open = mean(eyeDetection == 0)
  )

# Plot the proportion of time spent in each state over time intervals
ggplot(state_distribution, aes(x = as.numeric(time_interval))) +
  geom_line(aes(y = proportion_closed, color = "Closed")) +
  geom_line(aes(y = proportion_open, color = "Open")) +
  labs(x = "Time (seconds)", y = "Proportion", title = "Proportion of Eye States Over Time") +
  scale_color_manual(values = c("Closed" = "blue", "Open" = "red")) +
  theme_minimal()
```

This plot visualizes the EEG electrode intensities over time, with each line representing a different electrode. The vertical blue blocks indicate periods when the eyes are closed and red blocks indicate periods when the eyes are open. It seems that when the eyes are open, EEG signals decrease which may further prove our previous point that the brain is actively processing visual information from the environment. The changes in microvolt levels during periods of eye opening suggest a measurable impact of eye state on the brain's electrical activity.

Let's see if we can directly look at the distribution of EEG intensities and see how they related to eye status.

As there are a few extreme outliers in voltage we will use the `dplyr::filter` function to remove values outwith of 3750 to 50003. The function uses the `%in%` operator to check if each value of microvolts is within that range. The function also uses the `dplyr::mutate()` to change the type of the variable eyeDetection from numeric to a factor (R's categorical variable type).

```{r compare_distrib}
melt_train <- reshape2::melt(eeg_train, id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")

# filter huge outliers in voltage
filt_melt_train <- dplyr::filter(melt_train, microvolts %in% (3750:5000)) %>% dplyr::mutate(eyeDetection=as.factor(eyeDetection))

ggplot2::ggplot(filt_melt_train, ggplot2::aes(y=Electrode, x=microvolts, fill=eyeDetection)) + ggplot2::geom_boxplot()
```

Plots are great but sometimes so it is also useful to directly look at the summary statistics and how they related to eye status.
We will do this by grouping the data based on eye status and electrode before calculating the statistics using the convenient `dplyr::summarise` function.

```{r compare_summary_stats}
filt_melt_train %>% dplyr::group_by(eyeDetection, Electrode) %>% 
    dplyr::summarise(mean = mean(microvolts), median=median(microvolts), sd=sd(microvolts)) %>% 
    dplyr::arrange(Electrode)
```

**4** Based on these analyses are any electrodes consistently more intense or varied when eyes are open?



#### Time-Related Trends

As it looks like there may be a temporal pattern in the data we should investigate how it changes over time.  

First we will do a statistical test for stationarity:

```{r convert_to_tseries}
apply(eeg_train, 2, tseries::adf.test)
```


**5** What is stationarity?

Stationarity refers to a property of a time series in which its statistical properties, such as mean, variance, and autocorrelation, are constant over time. In other words, a stationary time series has no predictable patterns, trends, or seasonality, making its behavior more consistent over time.

Stationarity is an important concept in time series analysis and modeling because many statistical methods and models, such as ARIMA, require the data to be stationary. It can be either strict (or strong) stationarity and weak (or second-order) stationarity. In the first one, the entire joint distribution of the series remains unchanged over time, whereas for the latter one, only the first two moments (mean and variance) and the autocovariance function are invariant over time. Stationarity is usually tested through the Augmented Dickey-Fuller (ADF) Test, the Kwiatkowski-Phillips-Schmidt-Shin (KPSS) Test, and/or the Phillips-Perron (PP) Test.

**6** Why are we interested in stationarity? What do the results of these tests tell us? (ignoring the lack of multiple comparison correction...)

Stationarity is a crucial concept in time series analysis for several reasons, and the results of stationarity tests provide valuable insights into the nature and behavior of the data.



Then we may want to visually explore patterns of autocorrelation (previous values predict future ones) and cross-correlation (correlation across channels over time) using `forecast::ggAcf` function.

The ACF plot displays the cross- and auto-correlation values for different lags (i.e., time delayed versions of each electrode's voltage timeseries) in the dataset. 
It helps identify any significant correlations between channels and observations at different time points. 
Positive autocorrelation indicates that the increase in voltage observed in a given time-interval leads to a proportionate increase in the lagged time interval as well.
Negative autocorrelation indicates the opposite!


```{r correlation}
forecast::ggAcf(eeg_train %>% dplyr::select(-ds))
```


**7** Do any fields show signs of strong autocorrelation (diagonal plots)? Do any pairs of fields show signs of cross-correlation? Provide examples.


#### Frequency-Space 

We can also explore the data in frequency space by using a Fast Fourier Transform.  
After the FFT we can summarise the distributions of frequencies by their density across the power spectrum.
This will let us see if there any obvious patterns related to eye status in the overall frequency distributions.

```{r fft_open}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 0) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Open")
```

```{r fft_closed}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 1) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Closed")
```




**8** Do you see any differences between the power spectral densities for the two eye states? If so, describe them.



#### Independent Component Analysis

We may also wish to explore whether there are multiple sources of neuronal activity being picked up by the sensors.  
This can be achieved using a process known as independent component analysis (ICA) which decorrelates the channels and identifies the primary sources of signal within the decorrelated matrix.

```{r ica, warning=FALSE}
ica <- eegkit::eegica(eeg_train %>% dplyr::select(-eyeDetection, -ds), nc=3, method='fast', type='time')
mix <- dplyr::as_tibble(ica$M)
mix$eyeDetection <- eeg_train$eyeDetection
mix$ds <- eeg_train$ds

mix_melt <- reshape2::melt(mix, id.vars=c("eyeDetection", "ds"), variable.name = "Independent Component", value.name = "M")


ggplot2::ggplot(mix_melt, ggplot2::aes(x=ds, y=M, color=`Independent Component`)) + 
  ggplot2::geom_line() + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(mix_melt, eyeDetection==1), alpha=0.005) +
  ggplot2::scale_y_log10()
```



**9** Does this suggest eye opening relates to an independent component of activity across the electrodes?


### Eye Opening Prediction

Now that we've explored the data let's use a simple model to see how well we can predict eye status from the EEGs:

```{r xgboost}
# Convert the training and validation datasets to matrices
eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.numeric(eeg_train$eyeDetection) -1

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.numeric(eeg_validate$eyeDetection) -1

# Build the xgboost model
model <- xgboost(data = eeg_train_matrix, 
                 label = eeg_train_labels,
                 nrounds = 100,
                 max_depth = 4,
                 eta = 0.1,
                 objective = "binary:logistic")

print(model)
```



**10** Using the `caret` library (or any other library/model type you want such as a naive Bayes) fit another model to predict eye opening.

```{r model2}
# Load necessary libraries
library(e1071)
library(dplyr)
library(caret)

# Prepare the data
eeg_train_nb <- eeg_train %>% dplyr::select(-ds)
eeg_validate_nb <- eeg_validate %>% dplyr::select(-ds)

# Fit a Naive Bayes model
nb_model <- naiveBayes(eyeDetection ~ ., data = eeg_train_nb)

# Make predictions on the validation set
nb_preds <- predict(nb_model, eeg_validate_nb)

# Calculate accuracy
nb_accuracy <- mean(nb_preds == eeg_validate$eyeDetection)

# Print accuracy
print(paste("Naive Bayes Accuracy:", round(nb_accuracy, 2)))

# Confusion matrix and other metrics
nb_confusion_matrix <- confusionMatrix(nb_preds, eeg_validate$eyeDetection)
print(nb_confusion_matrix)
```
The overall accuracy of the Naive Bayes model is about 47.86%. This means that the model correctly predicted the eye status approximately 47.86% of the time.

Sensitivity (true positive rate) for eye opening is 20%. This means the model correctly identified 20% of the cases where the eyes were actually open. Moreover, Specificity (true negative rate) is 81.34%. This means the model correctly identified 81.34% of the cases where the eyes were actually closed. Other than that, balance accuracy is 50.67%, indicating that the model performs slightly better than random guessing overall.

The scores mean that the model struggles to correctly identify when the eyes are open and is better at identifying when the eyes are closed. Overall, the Naive Bayes model in this case is not effective at predicting eye opening from the EEG data.

**11** Using the best performing of the two models (on the validation dataset) calculate and report the test performance (filling in the code below):

First, we need to make predictions on the test data and evaluate performance metrics for both the models - XGBoost and Naive Bayes.

```{r test}
# Convert predicted labels to factor with appropriate levels
xgb_preds_factor <- factor(xgb_preds_binary, levels = levels(factor(eeg_test_labels)))

# Convert actual labels to factor with appropriate levels
eeg_test_labels_factor <- factor(eeg_test_labels, levels = levels(factor(xgb_preds_binary)))

# Calculate accuracy
xgb_accuracy <- mean(xgb_preds_factor == eeg_test_labels_factor)

# Print accuracy
print(paste("XGBoost Accuracy on Test Set:", round(xgb_accuracy, 2)))

# Confusion matrix and other metrics
xgb_confusion_matrix <- confusionMatrix(xgb_preds_factor, eeg_test_labels_factor)
print(xgb_confusion_matrix)
```
```{r test}
# Prepare the test data (assuming it's already prepared similar to validation data)
eeg_test_nb <- eeg_test %>% dplyr::select(-ds)

# Make predictions on the test set
nb_preds_test <- predict(nb_model, eeg_test_nb)

# Calculate accuracy
nb_accuracy_test <- mean(nb_preds_test == eeg_test$eyeDetection)

# Print accuracy
print(paste("Naive Bayes Accuracy on Test Set:", round(nb_accuracy_test, 2)))

# Confusion matrix and other metrics
nb_confusion_matrix_test <- confusionMatrix(nb_preds_test, eeg_test$eyeDetection)
print(nb_confusion_matrix_test)
```
We can see that XGBoost performs better than Naive Bayes. The accuracy from XGBoost is 84.35%. whereas that from Naive Bayes is 48.8%.



**12** Describe 2 possible alternative modeling approaches for prediction of eye opening from EEGs we discussed in the lecture but haven't explored in this notebook.


**13** What are 2 R libraries you could use to implement these approaches? (note: you don't actually have to implement them though!)

## Optional

**14** (Optional) As this is the last practical of the course - let me know how you would change future offerings of this course. This will not impact your marks!

- What worked and didn’t work for you (e.g., in terms of the practicals, tutorials, and lectures)?

- Was learning how to run the practicals on your own machines instead of a clean server that will disappear after the course worth the technical challenges?
 
- What would you add or remove from the course? 

- What was the main thing you will take away from this course?